\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{lscape}
\usepackage{todonotes}
%\usepackage{xcolor}
\usepackage{multirow}
\usepackage[table,xcdraw]{xcolor}
\usepackage{lscape}

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
\newcommand{\gp}[]{\textsc{gp }}

%\title{Prioritization for diagnosis through whole-genome sequencing of product of conception from idiopathic pregnancy losses }
\title{Paper Title}

\author[1+]{Silvia Buonaiuto}
\author[2+]{Imma Di Biase}
\author[3+]{Valentina Aleotti}
\author[4]{Adriano De Marino}
\author[1]{Gianluca Damaggio}
\author[5]{Marco Chierici}
\author[6]{Madhuri Pulijala}
\author[2]{Palmira D’Ambrosio} 
\author[2]{Gabriella Esposito}
\author[6]{Qasim Ayub}
\author[5]{Cesare Furlanello}
\author[7]{Nicole Soranzo}
\author[3]{Michele Rubini}
\author[4]{Antonio Capalbo}
\author[2]{Sebastiano Di Biase}
\author[1*]{Vincenza Colonna}
%\author[]{{\color{blue} nomi memorial exomes} }
\affil[1]{Affiliation, department, city, postcode, country}
\affil[2]{Affiliation, department, city, postcode, country}
\affil[*]{Correspondence: vincenza.colonna@igb.cnr.it (V.C.)}
\affil[+]{these authors contributed equally to this work}

%\keywords{Keyword1, Keyword2, Keyword3}

%\begin{abstract}

%PL impact on health - current diagnosis - It is estimated that up to 50\% of cases of PL do not have a clearly defined etiology.\cite{practice2012evaluation} - importance of diagnosis made on PoC: screening pre-IVF, genetic counselling for parents, research \\

%Current methods of diagnosis - What is missed: target on small-size genetic variants that causes PL  - How is this possible: diagnosis by whole genome sequencing,  not a routine, the protocol needs to be developed through large-scale projects  \\

%We present a collection of PL that complies to standards that we want to use for a pilot project - We use this collection to 1. draft a pipeline for samples prioritization before performing whole-genome sequencing 2. Estimate the size of the samples to be collected for large-scale studies\\

%We find: 20\% of samples goes to sequencing useful for determining sequencing costs and capabilities 

%\end{abstract}
\begin{document}

\flushbottom
\maketitle
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
%\include{intro}

\section*{Results}
To understand genetic susceptibility to miscarriage we studied the genome of forty-six miscarried embryos and nine women (not related to the forty-six embryos) experiencing spontaneous pregnancy termination. The embryos gestational age at pregnancy termination, calculated as the interval between the pregnancy termination date and the last menstruation date, ranges from 7.14 to 19.43 weeks (median is 10.3 weeks). Twenty-one embryos classifies as the product of recurrent miscarriages \cite{eshre2018eshre}. The mothers of the embryos are mostly of European origin (87\%) and their median age at the date of collection was 36.7$\pm$5.9 years, with no significant difference between first and recurrent cases (Figure \ref{fig:embryostats}, Mann–Whitney p-value=0.02). For the mothers of the embryos medical records report no major comorbidities. Folic acid was taken by 71\% of the mothers with no difference between first and recurrent cases (Figure \ref{fig:embryostats}, Chi-square p-value=0.96). Median body mass index and menarche age are comparable between first and recurrent cases, as well as comparable to a group of control women (Figure \ref{fig:embryostats}). Altogether, from the available medical records, we suppose that the recruited mothers of the embryos were in the range of healthy adult individuals.

{\color{blue}-- Antonio: description of women data set }

It is known from literature that roughly xx\% of the miscarriages in the first trimester are due to large chromosomal aneuploidies, such as trisomies or deletions of large chromosomal chunks \cite{goddijn2000genetic,zhang2009genetic}. In this study we want to focus on cases in which the genome is euploid, therefore the forty-six embryos were screened for chromosomal aneuploidies prior to whole-genome sequencing. We find that 32.6\% of samples were euploid and could be sequenced while 56.6\% of the embryos presented anuploidies (Figure \ref{fig:presequencing}).The most common aneuploidy in our data set is the trisomy of chromosome 22 (26.9\%), followed by trisomy of chromosome 16 (15.4\%). In particular, a first round of detection of aneuploidies on chromosomes 13, 15, 16, 18, 21, 22, X, and Y through Short Tandem Repeats analysis discarded 45.7\% of samples, %These types of repeats (tetra- or penta-nucleotide) are often expected to be found in heterozygosis, therefore triploidy is assumed when three alleles are found at several markers along a chromosome (complete) or part of it (partial). Similarly, uniparental disomy for a targeted region or chromosome is assumed when only one parental allele is amplified. 
and a subsequent analysis through comparative genomic hybridization and copy number variation detection form low-coverage sequencing discarded another 10.9\% of the samples. Finally, a number of embryos (10.9\%) dropped off the analysis due to low-quality DNA or maternal contamination. 
{\color{blue}--Antonio nine women kariotype}  

After ascertaining euploidy, the exome of the nine women was sequenced using {\color{blue}--Antonio sequencing} while the whole genome of ten embryos was sequenced using Illumina short-reads at 30X coverage. In the set of embryos genomes, we identified 11M single-nucleotide polymorphisms (SNPs) and 2M small insertions or deletions (indels), while from the exome data of the women we identified 1.7M SNPs and 276k indels.
%In the set of embryos genomes, we identified 11041,557 single-nucleotide polymorphisms (SNPs), 1980256 small insertions or deletions (indels), and XX copy number variants (CNVs). In the set of women we identified 1738895 SNPs and 275652 indels.


%Variant calling, phasing and annotation 
%Raw paired end sequencing data was mapped on the reference genome  GRCh38 using BWA (Li 2009) and variant calling was done using Freebayes (Garrison 2012), discovering on average 436k variants per sample with Phred-scaled quality scores >20. Information on functional annotations and Combined Annotation Dependent Depletion (CADD) for variable sites was obtained using Variant Effect Predictor (McLaren 2016). Phasing was done using Beagle 5.1 (Browning 2018) under standard parameters.  




%5493 genes in hgdp for exomes
%1323 discarded and 76% retained
\subsection*{Prioritization of genetic variants in coding genomic regions} 
We developed the \gp pipeline to prioritize putatively damaging genetic variants from sequencing data. \gp takes as input genomic variants information from cases and controls (including the per-individual allelic counts) in form of a vcf file and outputs a table of variants prioritized according to user-defined parameters. \gp uses functional annotations of genomic variants, information from publicly available sequence data of presumably healthy individuals, and, if available, knowledge of genes involved in the trait under study. \gp currently analyze coding regions and performs four filtering steps (Figure \ref{fig:pipeline}). The first filter (Filter I) retains variants based on: (i) a user-defined threshold of allele frequency in control populations; (ii) an overall impact on the gene product classified as moderate or high [REF]; (iii) the combined property of being putatively damaging (quantified by the CADD score [REF]) and located in genes intolerant to loss of function (determined by the pLI score [REF]). In addition it is possible to incorporate one or more user-defined lists of genes relevant to the trait under study, and take this into account when applying Filter I. Variants retained by Filter I (hits) are further filtered to control for false positives with Filters II and III. Finally, Filter IV excludes private variants with read depth outside the range found in non-private ones.   

%Filter II removes variants in genes with too many hits. Filter III controls for the chance of random occurrence of genes based on replicates of Filter I analysis in a large control population. In practice it removes all the genes that pass Filter I a user-defined-fraction of times across a user-defined-number of replicates. 

We applied the \gp pipeline to data from the high-coverage whole-genome and exome sequences of genomic DNA of the embryos the women (unrelated to embryos). For Filter I we set allele frequency <0.05\% in the 1000 Genomes [REF] and gnomAD[REF] reference populations, while the functional effect of the variant within the gene context was taken into account in two ways: either selecting for putatively deleterious variants (CADD score >90th percentile) in genes highly intolerant to loss of function (pLI score >0.9), or selecting for variants in genes known to be involved in early embryonic development. In particular for this last option we included five lists of genes, namely genes involved in embryo development (Gene Ontology GO:0009790), genes lethal during embryonic stages \cite{dawes2019gene}, essential for embryo development \cite{dawes2019gene}, genes discovered through the Deciphering Developmental Disorders project [REF DD], and a manually curated list of candidate genes known to be involved in miscarriages. We requested the variant to satisfy one or both these criteria: (i) be in a gene present in at least two of the five lists or (ii) have CADD score above the 90th percentile and be in a gene with pLI>0.9. Overall, filter I retained 1,038 and 728 variants (hits) in embryo and mothers, respectively.   

Filter II removed variants in genes with >5 hits. With few exceptions, we observed that the majority of genes (99th percentile) has less than five hits, even if there is no significant correlation between number of hits and gene length (Spearman r=XX p-value=xx, Figure \ref{fig:filters}A). Genes with >5 hits have more paralogues than genes with <5\% (???) and hits in gene with >5 hits are enriched for private variants  We concluded that variants found in gene with more than five hits are likely to be sequencing and alignment artifacts and therefore are discarded. 

For Filter III we used as control population 929 individuals from the Human Genome Diversity Project (HGDP, [REF]) from which we resampled 100 times ten individuals. On each resampled set we performed Filter I analysis and recorded the genes that were retained. Overall 5,488 unique genes were retained in controls with different frequencies in samples across replicates (Figure \ref{fig:filters}B ). When considering the 95th percentile, 1,531 genes are found >5\% of times across replicates, therefore hits within these genes were removed by filter III. Because the criteria used for Filter I were identical between embryos and mothers, and because the numerosity was very similar, we performed resampling only once and applied the outcome to both analyses.

Filter II and III, retained 447 and 359 hits in embryo and mothers, respectively of which 21\% in the embryos and 42\% in the women are private with respect to 1000 Genomes and gnomAD data sets. While in the embryos the read depth is comparable between private and non-privates variants, in women private variants have significantly lower coverage compared to non-private (Figure \ref{fig:filters}C-D, KS test p-value=XX, F test p-value=XXX). Therefore, to further control for possible artifacts due to scanty cverage, filter IV further removes hits which are private variants with read depth outside the range found in non-private ones.   

After all filters, \gp prioritize 439 and 357 unique variants in 399 and 320 genes, in the embryos and the women, respectively (Figures \ref{fig:resembryo} and \ref{fig:reswomen}). The genes prioritized in women are slightly enriched for genes belonging to mitochondrial translation initiation (R-HSA-5368286), elongation (R-HSA-5389840), and termination (R-HSA-5419276) pathways (p-value=0.0002, Benjamini-Hochberg adjusted p-value=0.0574, FDR=0.056). This enrichment is due to eight genes selected by \gp in six women due to missense mutations (Table \ref{tab:mito}).  --More about the eight genes %(\textit{MRPL15}, \textit{MRPL34}, \textit{MRPL37}, \textit{MRPS15}, \textit{MRPS21}, \textit{MRPS23}, \textit{MRPS28}, and \textit{OXA1L}) 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Properties and biological significance of the prioritized variants and genes} 
\subsubsection*{Mutations in \textit{STAG2},\textit{TLE4}, \textit{FMNL2}, and \textit{FRMPD3} in the embryos} 
In the embryos 4.1\% of the prioritized variants are stop gains/loss, frameshift indels, and variants that disrupt splicing sites, all classified as having high impact on the gene products, while missense mutations prevail among the variant with moderate effect (Figure \ref{fig:resembryo}A). The variants are hosted by 399 unique genes that code for 980 transcripts. 18.8\% of the prioritized genes were not in the lists of candidate genes used by \gp as input during the prioritization, demonstrating that \gp is robust to detection of genes never investigated before in relationship to the phenotype under study. Averages per embryos are 48.9$\pm$8.0 genomic variants in 47.8$\pm$7.7 genes coding for 113.5$\pm$24.6 transcripts (Figure \ref{fig:resembryo}B). For almost all genes \gp retains one variant per embryo, with few exception (five cases with two e and one with three variants per gene), as shown in Figure \ref{fig:resembryo}B, where the allele dosage and impact are also shown. %--Shared genes/variants 

%validation merigen? 
The male embryo AS030 carries one extremely rare T>G transversion (rs913664484, G frequency is 4.7e-05 in 42.7k individuals from gnomeAD [REF]) that changes one of the two basis at the 5' end of an the XX intron of the \textit{STAG2} gene (ENSG00000101972), altering a splice site (splice donor variant, high impact). \textit{STAG2} is on the X chromosome and codes for the cohesin subunit SA-2 \cite{cuadrado2020specialized}. Cohesins are ring-shaped protein complexes that brings into close proximity two different DNA molecules or two distant parts of the same DNA molecule \cite{mcnicoll2013cohesin}. In mouse, inactivation \textit{Stag2} causes early embryo lethality \cite{de2020essential}, and  mutations in \textit{STAG2} have been discovered in children with developmental disorders (REF DDD). Interestingly, only mildly-deleterious mutations have been found in males \cite{mullegama2019mutations}. 

The embryo AS036 carries an heterozygous missense G>C mutation (rs41307447) in the \textit{TLE4} (ENSG00000106829, synonym \textit{GRG-4}) gene that causes a substitution of a polar amino acid with another polar amino acid (Ser>Tre) in the XX exon of the gene, corresponding to a low complexity domain of the protein. The rs41307447 polymorphism is tolerated (SIFT score 0.18) and supposed to be benign (PolyPhen score 0.003), nevertheless the \textit{TLE4} gene is classified as highly intolerant to loss of function (pLI score 0.999) and the CADD score associated to rs41307447 is in the 99.8th percentile. 
\textit{TLE4} is a trascriptional repressor of the Groucho-family expressed in the embryonic stem cells where it repress naive pluripotency gene \cite{laing2015gro} and it is a direct transcriptional target of Notch \cite{menchero2019transitions}. \textit{TLE4} is also expressed in the extravillous trophoblasts \cite{meinhardt2014wnt} where it is part of the Wnt signaling pathway that promotes implantation, trophoblast invasion, and endo-metrialfunction \cite{sonderegger2010wnt}. Notably, a recent association study in a cohort of 750 cases of women experiencing recurrent miscarriages found a signal on chromosome 9 (rs7859844) in a genomic region physically interacting with \textit{TLE4} \cite{laisk2019genetic}, corroborating the hypothesis of association between variation in \textit{TLE4} and miscarriages. 

% AS093, AS090, AS087, AS065, AS036) 
Five embryos share one copy of an haplotype composed of two T alleles 4bp apart (rs866373641, rs750755379) in the \textit{FMNL2} gene (Table \ref{tab:fmnl2}). The two alleles exists at moderate-to high frequency in human populations (Table \ref{tab:fmnl2}) and are in perfect linkage disequilibrium (r\textsuperscript{2}=1) in the embryos, as well as in other cohorts (r\textsuperscript{2}= hgdp etc..). Women in this study .... 
We validated the association between the TT haplotype and phenotypic traits related to miscarriages in --UKBB 
In addition to the two mutations described above, the embryo AS065 has a missense mutation in phase (rs189416564, c.758C>G, p.Leu110Val, deleterious SIFT=0, probably damaging, PolyPhen = 0.969) 
\textit{FMNL2} codes for a formin-related protein and has 4 transcripts (splice variants). It is located on chromosome 2:152,335,174-152,649,826 forward strand. Formin-related proteins have been implicated in morphogenesis, cytokinesis, and cell polarity. \textit{FMNL2}, in particular, is expressed in the fetus in the cytoplasm of brain, spinal cord, and rectum\cite{lizio2015gateways}. \textit{FMNL2} is an elongation factor of actin filaments that drives cell migration by increasing the efficiency of lamellipodia protrusion \cite{block2012fmnl2} --more on FMNL2
    %It is implicated in cytoskeleton organization (GO:0007010), regulation of cell shape (GO:0008360), cellular component organization (GO:0016043), cell migration (GO:0016477), regulation of cell morphogenesis (GO:0022604), actin cytoskeleton organization (GO:0030036), cortical actin cytoskeleton organization (GO:0030866).

Two embryos (male AS065, female AS006) share a stop gain mutation in the X linked gene \textit{FRMPD3}, highly intolerant to loss of function (pLI = 0.91). Expressed in brain, ovary. thyroid (PROTEIN ATLAS). High expression of \textit{FRMPD3} in lion head goose with low production of eggs \cite{zhao2020genome} --more on FRMPD3 (where is the stop codon? which protein domain? )
%-- manually checked coverage FMNL2 FRMPD3 OK,  STAG2 DP basso EOMES check erik


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Results in the women} 
In the women 5.3\% of the prioritized variants are  high impact variants (stop gains/loss, frameshift indels, and variants that disrupt splicing sites) while missense mutations prevail among the variant with moderate effect (Figure \ref{fig:reswomen}A). The variants are hosted by 320 unique genes that code for 676 transcripts. 27\% of the prioritized genes were not in the lists of candidate genes used by \gp as input during the prioritization. Averages per women  are 42.5$\pm$10.9 genomic variants in 40.8$\pm$9.4 genes coding for 85.1$\pm$23.9 transcripts (Figure \ref{fig:reswomen}B). 
\noindent
\subsection*{to do}
-- comparisono with TRAPD\\ 
-- PADI6 PRDMA\\


\include{discussion}
\include{methods}
\bibliography{sample}
\section*{Acknowledgements (not compulsory)}
\section*{Author contributions statement}
\section*{Additional information}

\FloatBarrier
\include{figures}

\FloatBarrier
\include{tables}

\beginsupplement
\section*{Supplementary Information}
%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Supplementary Tables}
\include{supptables}
%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Supplementary Figures}
\include{suppfigure}


\end{document}